---
title: "w4 homework"
author: "Viktoria Pues"
date: "2023-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# week 4 homework 

## task 

1. Read in global gender inequality data
2. Join the global gender inequality index to spatial data of the World, creating a new column of difference in inequality between 2010 and 2019
3. Share it with the World on GitHub
4. Add you repository URL to the circulated spreadsheet

Tip the countrycode R package will be helpful!

Tip the gender inequality has changed in the last year, you will find what you need in the “All composite indices and components time series (1990-2021)” dataset, the metadata file beneath it will explain what the columns are.


### laod libraries 
```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(countrycode)
library(janitor)
library(here) 
library(tmap)
library(tmaptools)
library(sf)
library(here)
```

### read in the data 
global gender inequality data 
```{r}
gender_inequality <-read.csv(here("w4 data","HDR21-22_Composite_indices_complete_time_series.csv"), header = TRUE, sep = ",",  
                         encoding = "latin1")
```
spatial data world 
```{r}
spatial_data_world <- st_read("/Users/viktoriapues/Documents/casa/GIS/gis_w6-8/w4 data/World_Countries_Generalized/World_Countries_Generalized.shp")

print(spatial_data_world)

```
### join 
join on country ID column
creating a new column of difference in inequality between 2010 and 2019

```{r}
joined_data <- gender_inequality %>% select(iso3, country, hdi_2010, hdi_2019) %>% mutate(difference_2010_2019 = hdi_2019-hdi_2010) %>% 
  left_join(.,spatial_data_world,  by = c("country" = "COUNTRYAFF"))


```

